Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-07 ms.qmd
Rejig the first section of discussion to incoperate mroe species distributoon modelling

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011]. Here, we show that the relative variance ($lnCVR_{Q_{10}}$) in physiological rates for ectotherms is expected to decrease as temperatures increase, and this decrease is particularly pronouced in terrestrial ectotherms (~28%). Decreases in the relative variance of physiological rates are equal to or larger than the expected benefits from acclimation ($lnRR_{Q_{10_{acclimation}}}$) in marine and terrestrial ectotherms (*marine*: 16% acclimation vs. 15% reduction in relative variance; *terrestrial*: no acclimation vs 28% reduction in relative variance; @fig-1). Such effects are expected to result in a decrease in the variability in physiological rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide insight into the multifaceted ways in which climate warming will impact ectotherm populations, with important ecological and evolutionary implications that require further empirical and theoretical attention. We discuss our findings in the context of the current literature and highlight key areas for future research.

### ***Potential evolutionary consequences of reduced plasticity and variance in physiological rates across ectotherms***

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate]. Phenotypic plasticity can allow populations to persist in the face of climate extremes by matching phenotypes to fitness optima, although this is not always the case [@noble2019plastic; @radersma2020plasticity] and there may be limits and costs to plasticity that constrain its expression [@Dewitt1998; but see @Murren2015]. We show that plastic adjustments in physiological rates and changes in their relative variance differ across habitats. In freshwater habitats, acclimation responses will likely be beneficial considering that these effects are larger than reductions in relative variance in physiological rates. In contrast, acclimation responses of marine organisms coincide with a nearly equal reduction in relative variance in physiological rates, suggesting acclimation may not be as beneficial. Relative variance reductions are particularly pronounced in terrestrial ectotherms, and this coincides with a general inability to acclimate. Considering acute responses of animals acclimated to high temperatures are generally slightly elevated compared to cold-acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization is not likely to provide adaptive benefits under climate change. Together, these findings highlight the vulnerability of terrestrial ectotherms to climate change. Theoretical models that capture the percentage changes in relative variances and acclimation responses we observe here are needed fully appreciate the biological consequences such changes have on ecological and evolutionary dynamics. It may be expected that reductions in variance of this magnitude impact the 'opportunity for selection' by reducing the strength of selection and/or impacting underlying genetic variance exposed to selection. Both are predicted to impact the capacity of populations to evolve under climate change [@hoffmann2011climate; @urban2023and].

@@

@@

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were key traits most strongly impacted by increased temperature. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity in terrestrial ecosystems. However, data testing the impacts of changes in variance on populations are limited to very few taxa. We need creative experimentation across broader taxa that manipulate levels of variance to understand their ecological outcomes. Our meta-analytic findings can be integrated into theoretical models by permitting model parameters to capture biologically relevant changes in variation. Such predictive models can then be integrated with experimental approaches to better understand the consequences of changes in variation on ecological dynamics.

>>>>

Understanding acclimation capacity and how variation in physiological rates change across populations and species is important for predicting the ecological and evolutionary consequences of climate change [@Chevin2017; @Chevin2010; @Bolnick2011; @bush2016incorporating; @seebacher2023can]. Here, we show that the relative variance ($lnCVR_{Q_{10}}$) in physiological rates for ectotherms is expected to decrease to varying extents across terrestrial, marine and freshwater ecosystems as temperatures increase, with the decrease being particularly pronouced in terrestrial ectotherms (~28%). Such effects are expected to result in a decrease in the variability in physiological rates for marine and terrestrial ectotherms between ~4-13% under future climate change projections. Our results provide insight into the multifaceted ways in which climate warming will impact ectotherm populations that require further empirical and theoretical attention. We discuss our findings in the context of the current literature and highlight key areas for future research.

### ***Potential consequences of reduced plasticity and variance in physiological rates across ectotherms***

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate; @bush2016incorporating; @seebacher2023can]. For example, incorperating phenotypic plasticity and adaptation into species distribution models has been shown to dampen the predicted contraction of distributions in the face of climate change [e.g., @bush2016incorporating]. It addition, it may be expected that reductions in variance impact the 'opportunity for selection' by reducing the strength of selection and/or the genetic variance exposed to selection, impacting a population's capacity to evolve under climate change [@hoffmann2011climate; @urban2023and].

We show that plastic adjustments in physiological rates and changes in their relative variance differ across habitats. In freshwater habitats, acclimation responses will likely be beneficial and this coincides with relatively little reduction in the relative variance in physiological rates providing more raw material for selection to operate. In contrast, acclimation responses of marine organisms are associated with a nearly equal reduction in relative variance in physiological rates, which may trade-off with the beneficial effects that acclimation provides at high temperatures. In terrestrial ectotherms, relative variance reductions are particularly pronounced and this coincides with a general inability to acclimate. Considering acute physiological rates of animals acclimated to high temperatures are generally elevated compared to cold-acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization may not always provide adaptive benefits under climate change, particularly when considering changes in relative variance.

Our findings highlight the potential vulnerability of terrestrial and marine ectotherms to climate change. Our meta-analytic results provide opportunities to parameterise predictive models in ways that capture the percentage changes in relative variances and acclimation responses when predicting species distributions or modelling population resilience in the face of climate warnming. Our results also define the range of responses observed across diverse taxa (e.g., through prediction intervals) that also provide opportunities to incorperate realistic biological variation into the modelling process. Incorperating relavant variation in such change are needed fully appreciate the biological consequences such changes have on ecological and evolutionary dynamics (@bush2016incorporating).

@@

### ***Ecological consequences of reduced variation in physiological rates***
Variance reductions in metabolism and performance (e.g., sprint speed) were key traits most strongly impacted by increased temperature. Changes in variability in physiological rates, particularly traits governing energy demand, could have important consequences on the flow of energy within and between populations, communities, and ecosystems [@hendry2016key; @seebacher2023can; @barneche2021warming]. More variable populations, genetically and/or phenotypically, are predicted to be associated with broader niches, reduced intraspecific competition, increased growth rate, decreased vulnerability to environmental change and lower extinction risk [@forsman2015rethinking; @forsman2014effects; @Bolnick2011; @hendry2016key; @hart2016variation]. Maintaining intrapopulation variability in physiological rates in a warmer world may therefore be important for population resilience to climate change. For example, @kooijman1989population integrated individual variation in dynamic energy budget (DEB) models developed for *Daphnia magna*. They showed that greater between-individual variation in DEB parameters led to less extreme population fluctuations compared to scenarios with no variation [@kooijman1989population]. The differences in variance change across major habitat types suggest that terrestrial ectotherms will likely face greater challenges to adapting to climate change [@hoffmann2013upper; @Gunderson2015-hz; @Pottier2022], leading to greater rates of extinction and reduced productivity in terrestrial ecosystems. However, data testing the impacts of changes in variance on populations are limited to very few taxa. We need creative experimentation across broader taxa that manipulate levels of variance to understand their ecological outcomes. Our meta-analytic findings can be integrated into theoretical models by permitting model parameters to capture biologically relevant changes in variation. Such predictive models can then be integrated with experimental approaches to better understand the consequences of changes in variation on ecological dynamics.

@@
